Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-23 main_new.tex
updates

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task for
  modern marine management and science faced with limited resources to
  understand highly dynamic processes in the water-column. The
  combination of statistics and autonomous
  robotics provides opportunities for studying aspects related to
  experimental design for effective characterization of oceanographic
  phenomena. In this work we develop methods for efficient spatial
  sampling applied to the mapping of coastal processes by
  providing informative descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the joint excursions of random
  fields, and derive analytical expressions for the expected Bernoulli
  variance reduction in the excursion set of spatial temperature and
  salinity variables. We show how these can be applicable in static
  designs and sequential designs, building on myopic and look-ahead
  sampling strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches. Results from field
  deployments with an autonomous underwater vehicle are also presented,
  for a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage, but have limited resolution, are limited to
sensing the surface of the ocean, and is impacted by cloud cover. The
advent of robust mobile robotic platforms \cite{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular autonomous underwater vehicles (AUVs), have made
strides in advancing the state of sampling and consequently have made
robotics an integral part of ocean observation in large part by our own
work % driven by
% using statistical methods in design and data assimilation
\citep{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}.
Others include \cite{wikle2013modern} focusing on hierarchical
statistical models, \cite{sahu2008space} which studies spatio-temporal
models for sea surface temperature and salinity data, while
\cite{mellucci2018oceanic} looks at the statistical prediction of
features using an underwater glider. Common to these approaches is the
use of information-based methods to infer environmental sampling. A few
important characteristics should be mentioned, namely:

@@

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network, where robotic
  assets augment ship based measurements to increase coverage and
  resolution of phenomena of scientific interest.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Data-driven sampling under the sense-plan-act paradigm.]{\includegraphics[width =
    0.48\textwidth]{Figures/adaptive_cycle.pdf}\label{fig:sense-plan-act}}
  \caption{Ocean observation is moving away from single-ship sampling
    towards more autonomous and collaborative networked operations in
    order to resolve the numerous bio-geochemical processes and their
    interactions.}
\label{fig:envir}
\end{figure}

@@

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.85\textwidth]{Figures/pictures/c-updated.pdf}
\caption{Frontal patterns off of the Nidelva river, Trondheim, Norway.}
\label{fig:nidelven}
\end{figure}

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a range of scientific payloads to survey the
water column. Typically, they operate at $\sim 1-3$ m/s, can reach
depths from 100--6000 m with an in-water operation time capacity depending on
survey speed, payload sensors and mission design. By deploying AUVs
together with other resources one can augment information obtained
from either ocean models, remote-sensing data, or fixed-location
sensors on buoys. Traditional operations and surveys with AUVs are
usually limited to observations along fixed transects on a spatial
grid, pre-programmed by the human operator. By instead using on-board
algorithms to continuously evaluate, update, and refine future
sampling locations, one can make information gathering adaptive and
contextual to the survey requirements
\citep{das11b,fossuminformation,fossum18b}. The space of sampling
opportunities is still limited to the spatial grid, but now there is
much more flexibility because of adaptation. For instance, an AUV
could reconstruct or modify a survey line based on what temperatures
it is measuring by using embedded methods in decision-theoretic
planning \citep{py10,Rajan12,Rajan12b}.

@@

We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity.
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.eps}
  \caption{Density contour plots having different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. x-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean, $\mu_T=t_T=5^o C$ and
salinity $\mu_S=t_S=30$ mg/l. The displayed densities have unit
standard deviations for both temperature and salinity, but we also
study the effect of doubling the standard deviations.

@@

The optimal sequential design not only considers the best current AUV
grid node, but also what data gathered at this node would lead to for
future sampling at successive stages.  Let $d^{j,s}$ denote design
number $j$ at stage $s$ of a sequential survey. If this design is
selected, data $\by^{j,s}$ will be gathered. The optimal path
selection situation is then depicted in Figure \ref{fig:PathSelOpt},
\begin{figure}[h!]
\centering
\includegraphics[width=0.85\textwidth]{Figures/sequent_select.pdf}
\caption{Optimal sequential path selection.}\label{fig:PathSelOpt}
\end{figure}
where design choices are indicated by squares while data realizations are indicated by circles. 
%terms this optimal design is defined as
%\begin{equation}\label{opt_crit}
%    \bd^* = \mbox{argmin}_{d^{j,1}} \left\{ \int_{\by^{j,1}} \mbox{argmin}_{d^{j,2}} \left\{ \int_{\by^{j,2}} \ldots \pi(\by^{j,2}|\by^{j,1}) d\by^{j,2} \right\} \pi(\by^{j,1}) d \by^{j,1} \right\},
%\end{equation}
%where $\ldots$ represents the expected variance reduction in the ES under the continued optimal design, which will depend on the data at earlier stages. 
The mathematical expression for the optimal design then involves a
series of intermixed maximizations over designs and integrals over
data.  In practice the optimal solution is not tractable because of
the enormous growth over stages, see
e.g. \cite{sucar2015probabilistic} and \cite{powell2016perspectives}.
Instead, we outline heuristic strategies. In terms of notation, let
$\mathcal{Y}^{s-1} =\{d^{j,r}, \by^{j,r} ; r=1,\ldots,s-1 \}$ denote
the data gathered until stage $s-1$, using a selected design
$d^{j,r}$, $r=1,\ldots,s-1$.

@@

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.98\textwidth]{Figures/harald.jpg}
\caption{The commercially available Light Autonomous Underwater Vehicle (LAUV) platform (Harald) for upper water-
column exploration used in the experiment.}
\label{fig:lauv}
\end{figure}

>>>>

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for modern marine management and science faced with limited
  resources to understand highly dynamic processes in the
  water-column. The combination of statistics and autonomous robotics
  provides opportunities for studying aspects related to experimental
  design for effective characterization of oceanographic phenomena. We
  develop methods for efficient spatial sampling applied to the
  mapping of coastal processes by providing informative descriptions
  of spatial characteristics of ocean phenomena. Specifically, we
  define a design criterion based on reducing uncertainty in the joint
  excursions of random fields, and derive analytical expressions for
  the expected Bernoulli variance reduction in the excursion set of
  spatial temperature and salinity variables. We show how these can be
  applicable in static and sequential designs, building on myopic and
  look-ahead sampling strategies. We use simulations to study the
  properties and differences between state-of-the-art
  approaches. Results from field deployments with an autonomous
  underwater vehicle are also presented, for a case study mapping the
  boundary of a river plume. The results demonstrate the ability of
  the method to effectively inform and execute data-driven
  environmental sampling.

@@

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage, but have limited resolution, are limited to
sensing the surface of the ocean, and are impacted by cloud cover. The
advent of robust mobile robotic platforms \cite{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular autonomous underwater vehicles (AUVs), have
made strides in advancing the state of sampling and consequently have
made robotics an integral part of ocean observation; our own work
sheds has contributed to this effort % driven by
% using statistical methods in design and data assimilation
\citep{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}.
Others include \cite{wikle2013modern} focusing on hierarchical
statistical models, \cite{sahu2008space} which studies spatio-temporal
models for sea surface temperature and salinity data, while
\cite{mellucci2018oceanic} looks at the statistical prediction of
features using an underwater glider. Common to these approaches is the
use of information-based methods to infer environmental sampling. A few
important characteristics should be mentioned, namely:

@@

\begin{figure}[!t] 
  \centering 
  \subfigure[Illustration of an ocean sensing network, where robotic
  assets augment ship based measurements to increase coverage and
  resolution of phenomena of scientific interest.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Data-driven sampling under the sense-plan-act paradigm.]{\includegraphics[width =
    0.48\textwidth]{Figures/adaptive_cycle.pdf}\label{fig:sense-plan-act}}
  \caption{Ocean observation is moving away from single-ship sampling
    towards more autonomous and collaborative networked operations in
    order to resolve the numerous bio-geochemical processes and their
    interactions.}
\label{fig:envir}
\end{figure}

@@

\begin{figure}[!t] 
\centering 
\includegraphics[width=0.85\textwidth]{Figures/pictures/c-updated.pdf}
\caption{Frontal patterns off of the Nidelva river, Trondheim, Norway.}
\label{fig:nidelven}
\end{figure}

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a
range of scientific payloads to survey the water column. Typically,
they operate at $\sim 1-3$ m/s, can reach depths from 100--6000 m with
an in-water operation time capacity depending on survey speed, payload
sensors and mission design. By deploying AUVs together with other
resources one can augment information obtained from either ocean
models, remote-sensing data, or fixed-location sensors on
buoys. Traditional operations and surveys with AUVs are usually
limited to observations along fixed transects on a spatial grid,
pre-programmed by the human operator. By using on-board algorithms to
continuously evaluate, update, and refine future sampling locations,
one can make information gathering adaptive and contextual to the
survey requirements \citep{das11b,fossuminformation,fossum18b}. The
space of sampling opportunities while still limited to the spatial
grid, allows for adaptation. For instance, an AUV could reconstruct or
modify a survey line based on what temperatures it is measuring by
using embedded methods in decision-theoretic planning
\citep{py10,Rajan12,Rajan12b}.

@@

We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity.

\begin{figure}[!b] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.eps}
  \caption{Density contour plots having different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. x-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean, $\mu_T=t_T=5^o C$ and
salinity $\mu_S=t_S=30$ mg/l. The displayed densities have unit
standard deviations for both temperature and salinity, but we also
study the effect of doubling the standard deviations.

@@

The optimal sequential design not only considers the best current AUV
grid node, but also what data gathered at this node would lead to for
future sampling at successive stages.  Let $d^{j,s}$ denote design
number $j$ at stage $s$ of a sequential survey. If this design is
selected, data $\by^{j,s}$ will be gathered. The optimal path
selection situation is then depicted in Figure \ref{fig:PathSelOpt},
\begin{figure}[!t]
\centering
\includegraphics[width=0.85\textwidth]{Figures/sequent_select.pdf}
\caption{Optimal sequential path selection.}\label{fig:PathSelOpt}
\end{figure}
where design choices are indicated by squares while data realizations are indicated by circles. 
%terms this optimal design is defined as
%\begin{equation}\label{opt_crit}
%    \bd^* = \mbox{argmin}_{d^{j,1}} \left\{ \int_{\by^{j,1}} \mbox{argmin}_{d^{j,2}} \left\{ \int_{\by^{j,2}} \ldots \pi(\by^{j,2}|\by^{j,1}) d\by^{j,2} \right\} \pi(\by^{j,1}) d \by^{j,1} \right\},
%\end{equation}
%where $\ldots$ represents the expected variance reduction in the ES under the continued optimal design, which will depend on the data at earlier stages. 
The mathematical expression for the optimal design then involves a
series of intermixed maximizations over designs and integrals over
data.  In practice the optimal solution is not tractable because of
the enormous growth over stages, see
e.g. \cite{sucar2015probabilistic} and \cite{powell2016perspectives}.
Instead, we outline heuristic strategies. In terms of notation, let
$\mathcal{Y}^{s-1} =\{d^{j,r}, \by^{j,r} ; r=1,\ldots,s-1 \}$ denote
the data gathered until stage $s-1$, using a selected design
$d^{j,r}$, $r=1,\ldots,s-1$.

@@

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.98\textwidth]{Figures/harald.jpg}
\caption{The commercially available Light Autonomous Underwater
  Vehicle (LAUV) platform for upper water- column exploration used in
  the experiment. Image Courtesy LSTS, Univ. of Porto, Portugal.}
\label{fig:lauv}
\end{figure}
